---
title: "Aquaculture Analysis"
format: html
editor: visual
---

## Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(janitor)
library(estimatr)
library(performance)
library(jtools)
library(gt)
library(gtsummary)
library(interactions)
library(car)
```

## Data Wrangling

```{r}
#| message: false
#| warning: false
minnows <- read_csv(here::here("data", "Minnows_Aquaculture_Rennie19.csv")) %>%
  janitor::clean_names()

aquaculture <- minnows %>%
  # Adding fish sp names to ambiguous columns 
  rename(
    redbell_dace   = f182,
    finescale_dace = f183,
    fathead_minnow = f209,
    pearl_dace     = f214,
    slimy_sculpin  = f382
  ) %>%
  # Adding treatment column
  mutate(
    treat = case_when(lake == 375 ~ 1, TRUE ~ 0),
  # Add period column 
    period = case_when(
      year < 2003  ~ "Before",
      year <= 2007 ~ "During",
      TRUE         ~ "After"), 
  # Renaming season row names to be cleaner 
    season = case_when(season == "aSpring" ~ "Spring",
                       season == "zFall" ~ "Fall"),
  # Always order from Fall to Spring 
    season = fct_relevel(season, "Fall", "Spring")) %>%
  # Drop na values from minnow column 
  drop_na(minnows)

# Mean cpue per lake/year/season, sqrt-transformed
mean_cpue <- aquaculture %>%
  group_by(treat, year, season, period) %>%
  summarise(mean_cpue = mean(minnows), .groups = "drop") %>%
  mutate(
    sqrt_cpue = sqrt(mean_cpue))

# Prepare plot data
plot_data <- aquaculture %>%
  group_by(treat, year, season) %>%
  summarise(
    mean_cpue = mean(minnows),
    se_cpue   = sd(minnows) / sqrt(n()),
    .groups   = "drop"
  )
```

## Time Series Plot

```{r}
ggplot(plot_data, aes(x = year, y = mean_cpue, color = factor(treat))) +
  annotate("rect", xmin = 2003, xmax = 2007,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.5) +
  geom_line(aes(linetype = factor(treat))) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_cpue - se_cpue,
                    ymax = mean_cpue + se_cpue), width = 0.3) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "Year", y = "Average catch per unit effort",
       linetype = "", color = "") +
  theme_classic()
```

## BACI ANOVA

### Spring

```{r}
spring <- mean_cpue %>%
  filter(season == "Spring") %>%
  mutate(treat = as.factor(treat), period = as.factor(period))

mod_spring <- aov(sqrt_cpue ~ treat * period, data = spring)
summary(mod_spring)
TukeyHSD(mod_spring)
```

No significant terms for Spring ANOVA

`treat` no baseline differences in both lakes during spring

`period` spring minnow abundance did not change after introduction of aquaculture

`treat:period` The treated lake did not respond differently than the control lake across periods in spring. No aquaculture effect.

Spring Tukey HSD: Confirms no significant effect anywhere

### Fall

```{r}
fall <- mean_cpue %>%
  filter(season == "Fall") %>%
  mutate(treat = as.factor(treat), period = as.factor(period))

mod_fall <- aov(sqrt_cpue ~ treat * period, data = fall)
summary(mod_fall)
TukeyHSD(mod_fall)
```

All significant terms for Spring ANOVA:

`treat` significant differences in both lakes during fall

`period` fall minnow abundance changed over time in both lakes

`treat:period` lake 375 responded differently than lake 373 across periods in spring. Significant aquaculture effect.

Fall Tukey HSD:

`$treat` The control lake is 3.29409 sqrt CPUE units higher

`$period`: Post treatment was significantly higher than before; fall abundance spiked across both lakes in aquaculture years. No other significant term

`treat:period` significant terms:

`1:During-0: After`

`1:During-1: After`

`1:During-0: Before`

`1:During-1: Before`

`1:During-0: During`

## Causal Identification: Parallel Trends Test

Restricting to the Before period only, we test whether the two lakes were already trending differently prior to aquaculture. A significant `treat:year` interaction would indicate diverging pre-treatment trends, violating the parallel trends assumption.

```{r}
pre_only <- mean_cpue %>%
  filter(period == "Before")

mod_pt <- lm(sqrt_cpue ~ treat * year + season, data = pre_only)
summary(mod_pt)
```

```{r}
pre_only %>%
  ggplot(aes(x = year, y = mean_cpue,
             color = factor(treat), linetype = factor(treat))) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "Year", y = "Mean cpue (fish per net days)",
       color = "Lake", linetype = "Lake",
       title = "Pre-treatment trends (parallel trends check)") +
  theme_classic()
```

## Difference-in-Differences

### Before and During

```{r}
pre_aqua <- mean_cpue %>%
  filter(period != "After")

mod_did_pre <- lm(sqrt_cpue ~ treat * period + season, data = pre_aqua)

plot(mod_did_pre, which = 1)  # Residuals vs fitted
plot(mod_did_pre, which = 2)  # QQ-plot
plot(mod_did_pre, which = 4)  # Cook's distance
durbinWatsonTest(mod_did_pre) # Autocorrelation
summary(mod_did_pre)
```

### During and After

```{r}
post_aqua <- mean_cpue %>%
  filter(period != "Before")

mod_did_post <- lm(sqrt_cpue ~ treat * period + season, data = post_aqua)

plot(mod_did_post, which = 1)
plot(mod_did_post, which = 2)
plot(mod_did_post, which = 4)
durbinWatsonTest(mod_did_post)
summary(mod_did_post)
```

```{r}
post_aqua %>%
  ggplot(aes(x = year, y = mean_cpue,
             color = factor(treat), linetype = factor(treat))) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "Year", y = "Mean cpue (fish per net days)",
       color = "Lake", linetype = "Lake",
       title = "Post-treatment trends") +
  theme_classic()
```
