---
title: "Replicating the Aquaculture Impact Analysis on Minnow Communities"
author: "Kylie Newcomer, Joaquin Sandoval, and Grace Bryan"
date: "2026-03-06"
---

## Introduction

Freshwater aquaculture has the potential to alter lake ecosystems through nutrient inputs, changes in food availability, and shifts in fish community structure. Evaluating these ecological impacts is important for understanding how aquaculture operations influence freshwater biodiversity and ecosystem dynamics.

This report replicates part of the analysis presented in Rennie (2019), which examined ecosystem responses to an experimental rainbow trout aquaculture operation conducted at the Experimental Lakes Area in Ontario, Canada. The analysis focuses specifically on whether aquaculture activity influenced the abundance of minnows measured as catch-per-unit-effort (CPUE).

The study uses a Before-After Control-Impact (BACI) design comparing a treatment lake (Lake 375), where aquaculture was implemented, with a nearby control lake (Lake 373). Minnow CPUE is observed across three experimental periods: before aquaculture began, during the period of active aquaculture operations, and after the experiment ended.

While BACI designs are widely used in ecological impact studies, they rely on the assumption that the control site provides a valid counterfactual for the treated site. In this replication, we extend the standard BACI framework by explicitly testing the parallel trends assumption and by framing the analysis using a Difference-in-Differences (DiD) perspective. This approach allows us to evaluate whether treatment effects emerged during aquaculture operations and whether any effects persisted after the experiment concluded.

## Data Preparation

The dataset contains fish sampling observations collected from lakes within the Experimental Lakes Area. Each observation corresponds to a specific lake, year, and sampling season, and includes CPUE measurements for several fish species.

Two lakes are used in this analysis:

-   Lake 375, which served as the treatment lake where aquaculture operations were introduced

-   Lake 373, which served as a control lake

Minnow CPUE values are aggregated by lake, year, and season to produce an average abundance estimate for each sampling period. Because abundance data are often right-skewed, the CPUE values are square-root transformed prior to analysis to stabilize variance and better satisfy model assumptions.

The following code imports the dataset and constructs the variables used in the analysis.
=======
## Setup


```{r}
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(janitor)
library(estimatr)
library(performance)
library(jtools)
library(gt)
library(gtsummary)
library(interactions)
library(car)
```

### Data Wrangling

Key variables include:

-   treat - indicator for the treatment lake

-   period - before, during, and after aquaculture

-   season - spring or fall sampling
=======
## Data Wrangling

```{r}
#| message: false
#| warning: false
minnows <- read_csv(here::here("data", "Minnows_Aquaculture_Rennie19.csv")) %>%
  janitor::clean_names()

aquaculture <- minnows %>%
  # Adding fish sp names to ambiguous columns 
  rename(
    redbell_dace   = f182,
    finescale_dace = f183,
    fathead_minnow = f209,
    pearl_dace     = f214,
    slimy_sculpin  = f382
  ) %>%
  # Adding treatment column
  mutate(
    treat = case_when(lake == 375 ~ 1, TRUE ~ 0),
  # Add period column 
    period = case_when(
      year < 2003  ~ "Before",
      year <= 2007 ~ "During",
      TRUE         ~ "After"), 
  # Renaming season row names to be cleaner 
    season = case_when(season == "aSpring" ~ "Spring",
                       season == "zFall" ~ "Fall"),
  # Always order from Fall to Spring 
    season = fct_relevel(season, "Fall", "Spring")) %>%
  # Drop na values from minnow column 
  drop_na(minnows)

# Mean cpue per lake/year/season, sqrt-transformed
mean_cpue <- aquaculture %>%
  group_by(treat, year, season, period) %>%
  summarise(mean_cpue = mean(minnows), .groups = "drop") %>%
  mutate(
    sqrt_cpue = sqrt(mean_cpue))

# Prepare plot data
plot_data <- aquaculture %>%
  group_by(treat, year, season) %>%
  summarise(
    mean_cpue = mean(minnows),
    se_cpue   = sd(minnows) / sqrt(n()),
    .groups   = "drop"
  )
```

## Exploratory Time Series

Before estimating statistical models, it is useful to visualize how the minnow abundance changes over time in both lakes. This provides a preliminary view of whether the two lakes appear to respond differently during the aquaculture period.

```{r}

ggplot(plot_data, aes(x = year, y = mean_cpue, color = factor(treat))) +
  annotate("rect", xmin = 2003, xmax = 2007,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.5) +
  geom_line(aes(linetype = factor(treat))) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_cpue - se_cpue,
                    ymax = mean_cpue + se_cpue), width = 0.3) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "Year", y = "Average catch per unit effort",
       linetype = "", color = "") +
  theme_classic()

```

Figure 1 displays the time series of average minnow CPUE for the treatment and control lakes. The shaded region indicates the period during which aquaculture operations were active.

In spring sampling, the two lakes appear to follow broadly similar trajectories throughout the study period. In contrast, fall observations exhibit larger fluctuations in abundance, particularly in the treated lake during the aquaculture period. This increased variability suggests that any potential treatment effects may be more pronounced in the fall sampling, motivating the seasonal analyses presented below.

Visual inspection alone cannot establish causal effects, but these patterns help motivate the statistical models used in the following sections.

## BACI Analysis

A standard Before-After Control-Impact (BACI) analysis evaluates whether the treated site changes differently over time relative to a control site. Statistically, this is implemented using an interaction between treatment status and time period. The interaction term captures whether the change in the treated lake differs from the change observed in the control lake following the intervention.

In its simplest form, the BACI framework assumes that the control site provides a valid counterfactual for the treated site. If this assumption holds, a significant interaction term can be interpreted as evidence of an impact associated with the treatment.

### Spring BACI Model


```{r}
spring <- mean_cpue %>%
  filter(season == "Spring") %>%
  mutate(treat = as.factor(treat), period = as.factor(period))

mod_spring <- aov(sqrt_cpue ~ treat * period, data = spring)
summary(mod_spring)
TukeyHSD(mod_spring)
```


The results from the spring BACI model show no statistically significant effects for treatment, period, or their interaction. In other words, there is little evidence that minnow abundance in the treated lake changed differently from the control lake during the spring sampling period.

Post-hoc comparisons from the Tukey test also fail to detect significant differences across periods.

### Fall BACI Model

No significant terms for Spring ANOVA

`treat` no baseline differences in both lakes during spring

`period` spring minnow abundance did not change after introduction of aquaculture

`treat:period` The treated lake did not respond differently than the control lake across periods in spring. No aquaculture effect.

Spring Tukey HSD: Confirms no significant effect anywhere

### Fall


```{r}
fall <- mean_cpue %>%
  filter(season == "Fall") %>%
  mutate(treat = as.factor(treat), period = as.factor(period))

mod_fall <- aov(sqrt_cpue ~ treat * period, data = fall)
summary(mod_fall)
TukeyHSD(mod_fall)
```

The fall BACI model produces different results. In this case, treatment status, period, and their interaction are statistically significant. This suggests that minnow abundance in the treated lake changed differently over time compared to the control lake.

The post-hoc comparisons indicate that much of this difference appears during the aquaculture period, when CPUE in the treated lake increases relative to the control lake.

### Limitations of BACI

One limitation of the traditional BACI framework is that it assumes the control site provides a valid counterfactual for the treated site. In other words, the design assumes that the treated and control lakes would have followed similar trajectories in the absence of treatment.

If this assumption is violated, for example, if the lakes were already trending differently before aquaculture began, then the estimated BACI interaction could reflect those pre-existing differences rather than the effect of the intervention.

Difference-in-Differences approaches make this assumption explicit through the parallel trends assumption, which requires that treated and control units exhibit similar trends prior to treatment.

## Causal Identification: Parallel Trends Test

### Testing the Parallel Trends Assumption

To evaluate the parallel trends assumption, the dataset is restricted to the pre-treatment period. If the treated and control lakes followed similar trajectories prior to aquaculture operations, then there should be no systematic difference in trends between the two lakes during this period.

We test this by estimating a regression model that interacts treatment status with year using only pre-treatment observations. A non-significant interaction term provides evidence consistent with the parallel trends assumption.

Testing this assumption is particularly important in the present study because only one treated lake is available. With such limited replication, evidence that the lakes followed similar pre-treatment trajectories helps strengthen the causal interpretation of any post-treatment divergence.

All significant terms for Spring ANOVA:

`treat` significant differences in both lakes during fall

`period` fall minnow abundance changed over time in both lakes

`treat:period` lake 375 responded differently than lake 373 across periods in spring. Significant aquaculture effect.

Fall Tukey HSD:

`$treat` The control lake is 3.29409 sqrt CPUE units higher

`$period`: Post treatment was significantly higher than before; fall abundance spiked across both lakes in aquaculture years. No other significant term

`treat:period` significant terms:

`1:During-0: After`

`1:During-1: After`

`1:During-0: Before`

`1:During-1: Before`

`1:During-0: During`

## Causal Identification: Parallel Trends Test

Restricting to the Before period only, we test whether the two lakes were already trending differently prior to aquaculture. A significant `treat:year` interaction would indicate diverging pre-treatment trends, violating the parallel trends assumption.


```{r}
pre_only <- mean_cpue %>%
  filter(period == "Before")

mod_pt <- lm(sqrt_cpue ~ treat * factor(year) + season, data = pre_only)

mod_pt <- lm(sqrt_cpue ~ treat * year + season, data = pre_only)

summary(mod_pt)
```

```{r}
pre_only %>%
  ggplot(aes(x = year, y = mean_cpue,
             color = factor(treat), linetype = factor(treat))) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "Year", y = "Mean cpue (fish per net days)",
       color = "Lake", linetype = "Lake",
       title = "Pre-treatment trends (parallel trends check)") +
  theme_classic()

```


The coefficient of interest is the treatment-by-year interaction, which tests whether the treated lake was already trending differently before aquaculture began.

In this case, the interaction term is not statistically significant, suggesting that there is no strong evidence of diverging pre-treatment trends.

The time series shown in Figure 2 also provides a visual check of this assumption. The two lakes appear to follow broadly similar patterns prior to the start of aquaculture operations.

## Difference-in-Differences Analysis

After evaluating the parallel trends assumption, treatment effects are estimated using a Difference-in-Differences framework. In this approach, the interaction between treatment status and time period represents the estimated treatment effect, indicating whether the treated lake experienced a different change in CPUE compared to the control lake.

Because ecological responses may vary seasonally, models are estimated separately for spring and fall observations.

In addition, the post-treatment period is divided into two phases: the active aquaculture period and the post-aquaculture period. This allows the analysis to distinguish between immediate treatment effects and potential legacy effects after operations ended.

## Before and During
### Fall

```{r}
pre_aqua_fall <- mean_cpue %>%
  filter(period != "After",
         season == "Fall")

mod_did_pre_fall <- lm(sqrt_cpue ~ treat * period + factor(year), data = pre_aqua_fall)

# plot(mod_did_pre_fall, which = 1)  # Residuals vs fitted
# plot(mod_did_pre_fall, which = 2)  # QQ-plot
# plot(mod_did_pre_fall, which = 4)  # Cook's distance
# durbinWatsonTest(mod_did_pre_fall) # Autocorrelation
summary(mod_did_pre_fall)
```

The interaction coefficient in this model is positive and marginally significant, suggesting that minnow CPUE increased in the treated lake during aquaculture operations relative to the control lake.

 ### Spring
```{r}
pre_aqua_spring <- mean_cpue %>%
  filter(period != "After",
         season == "Spring")

mod_did_pre_spring <- lm(sqrt_cpue ~ treat * period + factor(year), data = pre_aqua_spring)

# plot(mod_did_pre_spring, which = 1)  # Residuals vs fitted
# plot(mod_did_pre_spring, which = 2)  # QQ-plot
# plot(mod_did_pre_spring, which = 4)  # Cook's distance
# durbinWatsonTest(mod_did_pre_spring) # Autocorrelation
summary(mod_did_pre_spring)
```
### Robustness Check
Adjusting the window
Add buffer period

## Before and After

```{r}
all_aqua_fall <- mean_cpue %>% 
  filter(season == "Fall")

mod_did_all_fall <- lm(sqrt_cpue ~ treat * period, data = all_aqua_fall)

# plot(mod_did_all_fall, which = 1)
# plot(mod_did_all_fall, which = 2)
# plot(mod_did_all_fall, which = 4)
# durbinWatsonTest(mod_did_all_fall)
summary(mod_did_all_fall)
```
```{r}
all_aqua_spring <- mean_cpue %>% 
  filter(season == "Spring")

mod_did_all_spring <- lm(sqrt_cpue ~ treat * period + factor(year), data = all_aqua_spring)

# plot(mod_did_all_spring, which = 1)
# plot(mod_did_all_spring, which = 2)
# plot(mod_did_all_spring, which = 4)
# durbinWatsonTest(mod_did_all_spring)
summary(mod_did_all_spring)
```

```{r}
# post_aqua %>%
#   ggplot(aes(x = year, y = mean_cpue,
#              color = factor(treat), linetype = factor(treat))) +
#   geom_line() +
#   geom_point() +
#   scale_color_manual(values = c("black", "red"),
#                      labels = c("Control", "Treated")) +
#   scale_linetype_manual(values = c("dashed", "solid"),
#                         labels = c("Control", "Treated")) +
#   facet_wrap(~ season, scales = "free_y") +
#   labs(x = "Year", y = "Mean cpue (fish per net days)",
#        color = "Lake", linetype = "Lake",
#        title = "Before and After treatment trends") +
#   theme_classic()
```

The interaction term in this model is not statistically significant. This suggests that the difference between the lakes becomes smaller once aquaculture operations end.

## Critical Evaluation

This analysis strengthens the traditional BACI framework by incorporating a Difference-in-Differences perspective and explicitly evaluating the assumptions required for causal interpretation.

Vanilla BACI designs test whether a treated site changed more than a control site following an intervention. However, this approach implicitly assumes that the control site represents a valid counterfactual trajectory. If the treated and control sites were already diverging prior to treatment, the interaction term could mistakenly attribute that divergence to the intervention.

The parallel trends test addresses this concern directly by examining whether the lakes were already trending differently before aquaculture began. The absence of a significant interaction provides some support for the validity of the control lakes as a counterfactual.

Another improvement in the analysis is the separation of the during-treatment and post-treatment periods. Traditional BACI analyses typically collapse these periods into a single "after" category. By estimating separate Difference-in-Differences models, this analysis distinguishes between immediate treatment effects and longer-term ecological responses.

However, the study still faces an important structural limitation: the experiment includes only one treated lake. With a single treated unit, statistical inference cannot fully separate treatment effects from idiosyncratic lake-level dynamics. As a result, the findings should be interpreted cautiously and viewed as evidence consistent with an aquaculture effect rather than definitive causal proof.

Despite this limitation, the combination of visualization, BACI modeling, parallel trends testing, and Difference-in-Differences analysis provides a coherent and transparent evaluation of the aquaculture experiment.

Overall, the results illustrate how combining BACI designs with a DiD framework can strengthen causal interpretation in ecological impact studies. While the analysis is limited by the presence of only a single treatment lake, the parallel trends test and the separation of during- and post-treatment periods provide a more transparent evaluation of how aquaculture activity may influence fish community dynamics.
=======
```

## Difference-in-Differences

### Before and During

```{r}
pre_aqua <- mean_cpue %>%
  filter(period != "After")

mod_did_pre <- lm(sqrt_cpue ~ treat * period + season, data = pre_aqua)

plot(mod_did_pre, which = 1)  # Residuals vs fitted
plot(mod_did_pre, which = 2)  # QQ-plot
plot(mod_did_pre, which = 4)  # Cook's distance
durbinWatsonTest(mod_did_pre) # Autocorrelation
summary(mod_did_pre)
```

### During and After

```{r}
post_aqua <- mean_cpue %>%
  filter(period != "Before")

mod_did_post <- lm(sqrt_cpue ~ treat * period + season, data = post_aqua)

plot(mod_did_post, which = 1)
plot(mod_did_post, which = 2)
plot(mod_did_post, which = 4)
durbinWatsonTest(mod_did_post)
summary(mod_did_post)
```

```{r}
post_aqua %>%
  ggplot(aes(x = year, y = mean_cpue,
             color = factor(treat), linetype = factor(treat))) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y") +
  labs(x = "Year", y = "Mean cpue (fish per net days)",
       color = "Lake", linetype = "Lake",
       title = "Post-treatment trends") +
  theme_classic()
```

