---
title: "Aquaculture Analysis"
format: html
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(estimatr)  
library(performance)
library(jtools)
library(gt)
library(gtsummary)
library(interactions)
library(ggridges)
library(ggbeeswarm)

minnows <- read_csv(here::here("data", "Minnows_Aquaculture_Rennie19.csv")) %>% 
  janitor::clean_names()

```

```{r}

aquaculture <- minnows %>% 
  rename(redbell_dace = f182,
         finescale_dace = f183,
         fathead_minnow = f209,
         pearl_dace = f214,
         slimy_sculpin = f382
         ) %>% 
  mutate(treat = case_when(
    lake == 375 ~ 1,
    TRUE ~ 0),
  is_post = case_when(
      year < 2003  ~ 0,
      year >= 2003 ~ 1
    ))  %>% 
  drop_na(minnows)
```

```{r}
### Mean cpue per lake/year/season, sqrt-transformed ###
mean_cpue <- aquaculture %>%
  group_by(treat, year, season, is_post) %>%
  summarise(mean_cpue = mean(minnows), .groups = "drop") %>%
  # Sqrt to normalize residuals 
  mutate(sqrt_cpue = sqrt(mean_cpue))

```

```{r}
# Prepare plot data 
plot_data <- aquaculture %>%
  group_by(treat, year, season) %>%
  summarise(
    mean_cpue = mean(minnows),
    se_cpue   = sd(minnows) / sqrt(n()),
    .groups  = "drop"
  )

season_labels <- c(aSpring = "Spring", zFall = "Fall") # Create clean seasons label

ggplot(plot_data, aes(x = year, y = mean_cpue, color = factor(treat))) +
  annotate("rect", xmin = 2003, xmax = 2007,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.5) +
  geom_line(aes(linetype = treat)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_cpue - se_cpue,
                    ymax = mean_cpue + se_cpue), width = 0.3) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y",
             labeller = labeller(season = season_labels)) +
  labs(x = "Year", 
       y = "Average catch per unit effort",
       linetype = "",
       color = "") +
  theme_classic()
```


```{r}
# BACI ANOVA for spring
spring <- mean_cpue %>% filter(season == "aSpring")%>% 
  mutate(treat = as.factor(treat),
         is_post = as.factor(is_post))
mod_spring <- aov(sqrt_cpue ~ treat * is_post, data = spring)
summary(mod_spring)
TukeyHSD(mod_spring)
```
No significant terms for Spring ANOVA 
`treat` no baseline differences in both lakes during spring 
`is_post` spring minnow abundance did not after introduction of aquaculture
`treat:is_post` The treated lake did not respond differently than the control lake across periods in spring. No aquaculture effect. 
Spring Tukey HSD: 
Confirms no significant effect anywhere
```{r}
# BACI ANOVA for fall
fall <- mean_cpue %>% filter(season == "zFall") %>% 
  mutate(treat = as.factor(treat),
         is_post = as.factor(is_post))
mod_fall <- aov(sqrt_cpue ~ treat * is_post, data = fall)
summary(mod_fall)
TukeyHSD(mod_fall)
```
All significant terms for Spring ANOVA 
`treat` significant differences in both lakes during fall 
`is_post` fall minnow abundance changed over time in both lakes 
`treat:is_post` lake 375 responded differently than lake 373 across periods in spring. Significant aquaculture effect. 
Fall Tukey HSD: 
`$treat` The control lake is 3.29409 sqrt CPUE units higher 
`$is_post`: Post treatment was significantly higher than before; fall abundance spiked across both lakes in aquaculture years. No other significant term 
`treat:is_post` all significant terms
```{r}
### Causal identification: parallel trends test
# Restrict to pre-treatment period only
pre_aqua <- mean_cpue  %>% 
  filter(is_post == 0) %>%
  mutate(season = factor(season, levels = c("zFall", "aSpring")))
# A significant lake:year interaction would violate parallel trends
mod_pt <- lm(sqrt_cpue ~ treat * is_post + season, data = pre_aqua)
summary(mod_pt)
# Visual parallel trends check
pre_aqua %>%
  ggplot(aes(x = year, y = mean_cpue,
             color = factor(treat),
             linetype = factor(treat))) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, size = 0.5) +
  scale_color_manual(values = c("black", "red"),
                     labels = c("Control", "Treated")) +
  scale_linetype_manual(values = c("dashed", "solid"),
                        labels = c("Control", "Treated")) +
  facet_wrap(~ season, scales = "free_y",
             labeller = labeller(season = season_labels)) +
  labs(x = "Year", y = "Mean cpue (fish per net days)",
       color = "Lake", linetype = "Lake",
       title = "Pre-treatment trends (parallel trends check)") +
  theme_classic()
```

### Difference-in-difference 
```{r}
library(car)
# DiD regression

mod_did <- lm(sqrt_cpue ~ treat * is_post + season, data = mean_cpue)

### Statistical assumption tests
# 1. Residuals vs fitted (homoscedasticity)
# which specifies type of diagnostic plot
plot(mod_did, which = 1) 
# 2. QQ-plot (normality)
plot(mod_did, which = 2)
# 3. Cook's distance (influential points)
plot(mod_did, which = 4)
# 4. Autocorrelation (Durbin-Watson)
durbinWatsonTest(mod_did)

summary(mod_did)
```
From `mod_did` summary : 
`treat` 0.2655 , p = 0.829. No significant baseline difference between lakes before treatment. 
`is_post` 0.3466, p = 0.7347. No significant change in CPUE after treatment
`seasonzFall` **3.1815, p =  0.0001**  significantly higher CPUE than spring regardless of treatment 
`treat:is_post` 2.6080, p = 0.100356. There is a 2.6 increase in CPUE of the treated lake in the time period after treatment began

R2 = 0.3191 - 32% of variation of minnow CPUE across lake/year/season is explained by model. 
F = 6.986, p = 9.292e-05. Model significant - explains more variance than just using mean CPUE as prediction. 




